The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2006-09-23 trunk/vcd/paper/Strucplot/JSS.Rnw
several fixes in JSS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@575 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The remainder of the paper is organized as follows. In
Section \ref{sec:strucplot}, we give an overview of the strucplot
framework, describing the hierarchy of the main components and the
basic functionality. 
In Section \ref{sec:shading}, we demonstrate how
(residual-based) shadings support the visualization of log-linear models and
the results of independence tests. Also, we explain step-by-step how
the concepts of generating functions and graphical appearance control
functions can be combined to provide a flexible customization of
complex graphical displays as created by the strucplot framework.
Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
and spacing features, respectively, before Section \ref{sec:conclusion}
concludes the work.

@@

<<structable,results=verbatim>>=
(hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
@

@@

\noindent All these operations scale linearly with the amount of
created viewports. For a $d$-dimensional table with $k_i$
levels, $i=1 \dots d$, the total number of needed viewports
$T_d$ can roughly be estimated as:

\begin{equation}
  \label{eq:numbervp}
  T_d = \sum_{i=1}_d \prod_{j \le i} k_j
\end{equation}

\noindent Assuming an equal number for all dimensions, we obtain:

\begin{equation}
  \label{eq:equalvp}
  T_d = \sum_{i=1}_d k^i = \frac{k(k^d-1)}{k-1}
\end{equation}

\noindent and so the time complexity for drawing a strucplot display is of order $O(k^d)$.

@@

@@

\begin{description}
\item[\texttt{Arthritis}] Data from \cite{vcd:Koch+Edwards:1988} from a 
  double-blind clinical trial
  investigating a new treatment for rheumatoid arthritis. Package:
  \pkg{vcd}.
\item[\texttt{Bundesliga}] Results from the first German soccer league in the
  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
  Zeileis). Package: \pkg{vcd}.
\item[\texttt{HairEyeColor}] Distribution of hair and eye color and 
  gender in 592 statistics
  students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
  Package: \pkg{datasets} (included in base \proglang{R}).
\item[\texttt{PreSex}]  Data from \cite{vcd:thornes+collard:1979} reported in 
  \cite{vcd:gilbert:1981}, on pre- and extra-marital sex and divorce. Package:
  \pkg{vcd}.
\item[\texttt{Titanic}] Information on the fate of passengers on
  the fatal maiden voyage of the ocean liner `Titanic', summarized
  according to economic status (class), gender (`Sex'), age and
  survival.  Source: \cite{vcd:dawson:1995} (Data 
  originally collected by the British Board of Trade
  in their investigation of the sinking). 
  Package: \pkg{datasets} (included in base \proglang{R}).
\item[\texttt{UCBAdmissions}] 
  Aggregate data on applicants to graduate school at Berkeley for
  the six largest departments in 1973 classified by admission and
  gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. 
  Package: \pkg{datasets} (included in base \proglang{R}).
\end{description}

@@

\item Mosaic plot for the \data{Arthiritis} data in Figure \ref{fig:arthritis}:
\begin{Sinput}
mosaic(art, gp = shading_max, split_vertical = TRUE)
\end{Sinput}

>>>>

The remainder of the paper is organized as follows. In
Section \ref{sec:strucplot}, we give an overview of the strucplot
framework, describing the hierarchy of the main components and the
basic functionality. 
In Section \ref{sec:shading}, we demonstrate how
(residual-based) shadings support the visualization of log-linear models and
the results of independence tests. Also, we explain step-by-step how
the concepts of generating functions and graphical appearance control
functions can be combined to provide a flexible customization of
complex graphical displays as created by the strucplot framework.
Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
and spacing features, respectively. Section \ref{sec:example}
exemplifies the framework in the analysis of a four-way data set, 
before Section \ref{sec:conclusion} concludes the work.

@@

<<structable,echo=TRUE,results=verbatim>>=
(hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
@

@@

\noindent All these operations scale linearly with the amount of
created viewports. For a $d$-dimensional table with $k_i$
levels, $i=1 \dots d$, the total number of needed viewports
$T_d$ can roughly be estimated as

\begin{equation}
  \label{eq:numbervp}
  T_d = k_1 + k_1k_2 + \dots + k_1 \dots k_d = \sum_{i=1}^d \prod_{j \le i} k_j
\end{equation}

\noindent since we first push the $k_1$ viewports for the levels of the
first dimension, then, for \emph{each} of those, the $k_2$ levels of
the second dimension, etc. Assuming an equal number of levels $k$
for all dimensions, we obtain:

\begin{equation}
  \label{eq:equalvp}
  T_d = \sum_{i=1}^d k^i = \frac{k(k^d-1)}{k-1}
\end{equation}

\noindent and so the time complexity for drawing a strucplot display is of order $k^d$.

@@

\section{Example: the Ovarian Cancer Data}
\label{sec:example}

In the following, we demonstrate some of the described techniques in
analyzing the Ovarian Cancer data set.
The data origins from \citep{vcd:obel:1975} and is 
about a retrospective study of ovary cancer
carried out in 1973.  Information was obtained from 299 women, who
were operated for ovary cancer 10 years before. The data consists of
four binary variables: the `stage' of the cancer at the time of
operation (levels: early, advanced), the type of `operation' performed
(radical, limited), the `survival' status after 10 years (yes, no),
and `xray' indicating whether X-ray treatment was received (yes,
no).

The dataset in \pkg{vcd} comes pretabulated in a data frame, so we
first create the four-way table:

<<oc1>>=
tab <- xtabs(Freq ~ stage + operation + xray + survival, data = OvaryCancer)
@

\noindent A `flattened' textual representation can be obtained using \codefun{structable}:

<<oc2,echo=TRUE,results=verbatim>>=
structable(survival ~ ., data = tab)
@

\noindent A first overview can be obtained using a pairs plot (Figure \ref{fig:ocpairs}):

<<oc3>>=
dpa <- list(var_offset = 1.2, rot = -30, just_leveltext= "left")
pairs(tab, diag_panel_args = dpa)
@

\begin{figure}[h]
\begin{center}
<<ocpairs,echo=FALSE,fig=TRUE>>=
<<oc3>>
@ 
\caption{Pairs plot for the \data{OvaryCancer} data showing mosaic displays for all pairwise distributions and bar plots for all marginal distributions.}
\label{fig:ocpairs}
\end{center}
\end{figure}

\noindent The pairs plot, by default, creates mosaic displays for all pairwise 
variable combinations, and bar plots in the diagonal to visualize the absolute 
frequencies of the variables. The \texttt{var\_offset} argument modifies the offset of the 
(centered) variable names to avoid overlap with the bars. 
Additionally, we use the \texttt{rot} and the
\texttt{just\_leveltext} arguments to rotate the level names, again to avoid their overlap.
First, we consider the marginal distributions. The study design involved 
(nearly) the same number of survived (150) and deceased (149) patients. Similarly balanced,
158 cases were in an advanced and 141 in an early stage. Most patients (251, 84\%) 
were treated with a radical operation, and 186 (62\%) got X-ray treatment. Next, we inspect
the two-way interaction of the influencing factors (`stage', `operation', and `xray'): the
corresponding mosaics exhibit symmetric, regular shapes with aligned tiles, 
which indicate no marginal interaction between these variables. 
The same is true for the interactions of `survival'
with `operation' and `xray', respectively. Only the stage seems to influence
survival: here, the tiles are ``shifted''.

A different view on the data can be obtained using a doubledecker plot, explaining
survival by the three other factors (Figure \ref{fig:ocdoubledecker}):

<<oc4>>=
doubledecker(survival ~ stage + operation + xray, data = tab)
@

\begin{figure}[h]
\begin{center}
<<ocdoubledecker,echo=FALSE,fig=TRUE>>=
<<oc4>>
@ 
\caption{Doubledecker plot for the \data{OvaryCancer} data showing the
 conditional distribution of `xray', given `operation', given `stage', and with
 `survival' highlighted.}
\label{fig:ocdoubledecker}
\end{center}
\end{figure}

\noindent From a technical point of view, the display is constructed as a mosaic plot
showing the conditional distribution of `survival', given `xray', given `operation', 
given `stage', with vertical splits for the conditioning variables and horizontal ones for
`survival'. Additionally, there is zero space between the tiles of the last dimension and
a binary shading is used for survived and deceased patients.
Conceptually, this plot is interpreted as a mosaic plot of just the influencing
variables, with `survival' highlighted in the tiles. Thus, the plot really shows
the influence of the ``explanatory'' variables on `survival'. Clearly, the survival
rate is higher among patients in an early stage, but neither radical operation 
nor X-ray treatment seem to improve the situation. From this exploratory
phase, it even seems that for patients in an
early stage with a limited operation, no X-ray treatment would be better, 
and that for patients in an advanced stage, limited in contrast to radical 
operation would improve the chance of survival. %'

To visualize inference results, we can make use of residual-based shadings. First, we
again use the pairs plot display, this time using color to bring out departures from
the hypothesized marginal independence models (Figure \ref{fig:ocpairsshaded}):

<<oc5>>=
pairs(tab, diag_panel_args = dpa, gp = shading_max)
@

\begin{figure}[h]
\begin{center}
<<ocpairsshaded,echo=FALSE,fig=TRUE>>=
<<oc5>>
@ 
\caption{Pairs plot for the \data{OvaryCancer} data showing mosaic displays for all pairwise distributions and bar plots for all marginal distributions. The colored cells show the
violation of the independence hypothesis for `stage' and `survival'.}
\label{fig:ocpairsshaded}
\end{center}
\end{figure}

\noindent Only the mosaics showing the interaction of `stage' and `survival' have 
colored cells, indicating that the survival rate of patients in an early stage is
higher than would be expected under (marginal) independence. The hypotheses of
independence are not rejected for all other marginal interactions. Next, we investigate
log-linear models for the four-way table. The default model used by \codefun{mosaic} is
the total independence model, visualized in Figure \ref{fig:ocmosaictotal}:

<<oc6>>=
mosaic(tab, shade = TRUE)
@

\begin{figure}[h]
\begin{center}
<<ocmosaictotal,echo=FALSE,fig=TRUE>>=
<<oc6>>
@ 
\caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
for the (clearly rejected) default hypothesis of total independence of all variables.}
\label{fig:ocmosaictotal}
\end{center}
\end{figure}

\noindent The model is clearly rejected (P-value: 0.000). From the exploratory phase of
our analysis, we (only) suspect `stage' to be influential on the survival rate. A 
corresponding hypothesis is that `survival' be independent of `xray' and `operation', given
`stage'. The model is specified using the \texttt{expected} argument, 
either using the \codefun{loglin} interface or 
the \codefun{loglm} formula interface (the resulting mosaic plot is shown in Figure
\ref{fig:ocmosaicstage}):

<<oc7>>=
mosaic(tab, expected = ~ (survival + operation * xray) * stage)
@

\begin{figure}[h]
\begin{center}
<<ocmosaicstage,echo=FALSE,fig=TRUE>>=
<<oc7>>
@ 
\caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
for the hypothesis of `survival' being independent of `xray' and `operation', given
`stage'. The hypothesis is not rejected.}
\label{fig:ocmosaicstage}
\end{center}
\end{figure}

\noindent Thus, based on this data, only pre-diagnosis seems to matter in ovarian
cancer therapy.

@@

\begin{description}
\item[\texttt{Arthritis}] Data from \cite{vcd:Koch+Edwards:1988} from a 
  double-blind clinical trial
  investigating a new treatment for rheumatoid arthritis. Package:
  \pkg{vcd}.
\item[\texttt{Bundesliga}] Results from the first German soccer league in the
  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
  Zeileis). Package: \pkg{vcd}.
\item[\texttt{HairEyeColor}] Distribution of hair and eye color and 
  gender in 592 statistics
  students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
  Package: \pkg{datasets} (included in base \proglang{R}).
\item[\texttt{OvaryCancer}]  Data about a retrospective study of ovary cancer
     carried out in 1973.  Information was obtained from 299 women, who
     were operated for ovary cancer 10 years before. Source:
     \cite{vcd:obel:1975}. Package: \pkg{vcd}.
\item[\texttt{PreSex}]  Data from \cite{vcd:thornes+collard:1979} reported in 
  \cite{vcd:gilbert:1981}, on pre- and extra-marital sex and divorce. Package:
  \pkg{vcd}.
\item[\texttt{Titanic}] Information on the fate of passengers on
  the fatal maiden voyage of the ocean liner `Titanic', summarized
  according to economic status (class), gender (`Sex'), age and
  survival.  Source: \cite{vcd:dawson:1995} (Data 
  originally collected by the British Board of Trade
  in their investigation of the sinking). 
  Package: \pkg{datasets} (included in base \proglang{R}).
\item[\texttt{UCBAdmissions}] 
  Aggregate data on applicants to graduate school at Berkeley for
  the six largest departments in 1973 classified by admission and
  gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. 
  Package: \pkg{datasets} (included in base \proglang{R}).
\end{description}

@@

\item Mosaic plot for the \data{Arthritis} data in Figure \ref{fig:arthritis}:
\begin{Sinput}
mosaic(art, gp = shading_max, split_vertical = TRUE)
\end{Sinput}
